When Risk Hits the Road: Does Danger Drive a Return?

### 2024-12-03 paper/paper.qmd
Add reference section
Finished the whole paper

---
title: "When Risk Hits the Road: Does Danger Drive a Return?"
subtitle: "Risk Peaks in the Morning on Dark, Snow-Packed Roads and in High Urban Density Areas"
author: 
  - Yingke He
thanks: "Code and data are available at: [https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main](https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main)."
date: today
date-format: long
abstract: "This study examines motor vehicle theft and traffic collisions in Toronto, focusing on spatial and temporal patterns,and environmental factors. A composite risk score model is used and reveals that collisions occur more frequently under unfavorable road conditions, such as poor lighting and wet or icy surfaces, and in areas with inadequate traffic control measures. These findings emphasize the need for targeted interventions, including enhanced road infrastructure and lighting in collision-prone zones. Such measures aim to improve safety and security for motorbike riders and owners, contributing to safer urban mobility in Toronto."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The structure of the paper is organized as follows: following @sec-intro, @sec-data outlines the data collection and preprocessing procedures, along with a detailed description of the outcome variable and the predictor variables used in the analysis. @sec-model introduces the logistic regression models applied to estimate the probability of collision, as well as the method used to derive the theft index and combine these probabilities into a composite risk score. @sec-result then presents the main findings, including insights into how different factors contribute to the risks of owning and riding a motorbike. Finally, @sec-discussion interprets the results, highlighting significant trends and implications for motorbike risk assessment, and concludes with a discussion on the limitations of the analysis and future research directions.

@@

Details about the data cleaning process and the criteria for variable selection are provided in Appendix A.

@@

The Theft from Motor Vehicle dataset contains all reported occurrences of thefts from vehicles, categorized by reported date. These offences are classified based on the value of the stolen items, distinguishing between theft under and theft over thresholds. Each occurrence number may include multiple rows, representing the various offences associated with a single event. The dataset excludes "unfounded" occurrences, adhering to Statistics Canada’s definition that these events were determined not to have occurred or been attempted. Like the KSI dataset, this dataset includes fields for both the old and new neighborhood structures, enabling comprehensive geographic analyses of theft trends.

@@

@fig-risk-index-histogram shows the distribution of the Risk Index across all observations in the dataset. The Risk Index shows a unimodal distribution, skewed slightly to the left, with the majority of values concentrated between 0.45 and 0.65. This indicates that most motorbike-related risks fall within a moderate range. The density curve overlay reveals a smooth progression in risk levels, with the peak occurring around a Risk Index value of 0.55, suggesting that this is the most common level of composite risk. The left tail, representing lower risk levels, is relatively small, while the right tail, corresponding to higher risks, extends further, indicating the presence of a smaller number of high-risk cases.

The skewness and spread of the distribution highlight the variability in the combined risks of theft and collision. The extended tail on the higher end of the Risk Index suggests that certain environmental or situational factors disproportionately increase the risks in specific cases. This insight can inform targeted interventions, focusing on the outliers with high Risk Index values to mitigate the most critical risks.

@@

Building on the distribution of the Risk Index in @fig-risk-index-histogram, @fig-risk-index-boxplot highlights the distribution of the Risk Index across three defined categories: High Risk, Low Risk and Moderate Risk. Each category represents a grouping of neighborhoods based on their average Risk Index values. The plot reveals distinct differences in the distribution of the Risk Index between these categories.

@@

The **Hour of the Day** variable is used as a predictor to analyze temporal trends in motorbike theft occurrences. This variable helps identify whether certain hours are associated with elevated or reduced theft risks. Contributing factors may include decreased monitoring during nighttime hours, increased activity in high-risk areas during specific times, or patterns related to commuter schedules. @fig-hour_summary outlines the distribution of bicycle theft incidents across different hours of the day in a 24-hour format. This temporal variable helps the model capture time-dependent patterns in theft occurrences, identifying periods of heightened risk, such as late morning to early afternoon (10:00 to 15:00), and lower-risk periods, such as early morning hours (02:00 to 06:00).By including this variable, the theft index model utilizes the observed trend of thefts being more frequent during morning and early afternoon hours (7:00 to 13:00) while being significantly lower during the early morning hours (02:00 to 06:00). This temporal variability highlights specific periods of increased theft risk, enabling the model to capture these patterns effectively. Understanding this trend allows for better predictions of theft dynamics and offers valuable insights for designing targeted interventions and optimizing resource allocation based on the time of day.

@@

The Collision Probability model, on the other hand, examines situational and environmental conditions affecting the likelihood of collisions. Variables such as Neighborhood ID, Road Surface Condition, Lighting Condition, Traffic Control, Time of Day, and Visibility offer insights into the contextual factors that contribute to traffic incidents, capturing the dynamic interaction between environmental factors and human behavior.

@@

The **Traffic Control** variable categorizes the type of traffic management present at the site of each incident, such as "No Control," "Traffic Signal," or "Stop Sign." This variable provides insight into how various traffic control measures influence the likelihood of motorbike collisions.

@@

Detailed model diagnostics, variable descriptions, and performance metrics are included in [Appendix @sec-model-details].

@@

This modeling approach is justified by its ability to adapt to the data’s constraints while maintaining statistical rigor and interpretability. The sub-index method for theft risk is tailored to the dataset’s characteristics, avoiding assumptions about unobserved cases, and the use of logistic regression for collision risk ensures robust and reliable predictions. Together, these components form a comprehensive and practical framework for evaluating motorbike ownership and usage risks, addressing both immediate safety concerns and long-term theft risks.

@@

where P represents the predicted probability of collision from the logistic regression model plotted against the continuous predictors. These plots are intended to visualize the relationship between each predictor and the logit of the outcome variable, providing insight into whether the relationship is approximately linear.

@@

Despite this overlap, both predictors are retained in the model because they provide distinct and meaningful contributions to understanding collision risk. Visibility Conditions directly reflects environmental factors like fog, heavy rain, or low light, which impair drivers' ability to see hazards. Conversely, Road Conditions\` account for the physical state of the driving surface, such as wet, icy, or damaged roads, which influence vehicle handling and stopping distance. Together, these variables capture complementary aspects of collision risk, ensuring that the model provides a comprehensive assessment.

@@

The year 2020 coincides with the beginning of the COVID-19 pandemic, which had a profound impact on urban mobility patterns. Lockdowns, social distancing measures, and reduced economic activities likely led to significant decreases in traffic volumes and changes in behavior related to motor vehicle usage [@buehler2021covid]. However, the Risk Index remains relatively consistent throughout the year, with only minor fluctuations across months. This relative stability may suggest that while pandemic-related restrictions likely influence traffic and mobility patterns, these changes did not significantly impact the overall Risk Index for the year. The consistent trends could reflect a balance between reductions in traffic collisions due to decreased activity during lockdown periods and a stable baseline for theft and other risk factors.

@@

This section examines how factors such as road surface conditions, lighting conditions, and traffic control types contribute to changes in the Risk Index. Categorizing the data based on these factors reveals specific conditions under which the likelihood of collisions or theft increases.

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing insight into how these factors may contribute to the overall risk.
```{r}
#| label: fig-risk-distribution-roadcondition
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Road Surface Conditions"
#| fig-pos: H
#| fig-width: 8.5
#| fig-height: 4

@@

@fig-risk-distribution-roadcondition reveals notable differences in the Risk Index distribution across various road surface conditions. Wet and icy conditions, in particular, show higher variability in the Risk Index, suggesting that these surfaces may lead to more unpredictable and elevated risk levels. Dry conditions, on the other hand, generally exhibit a lower and more consistent risk. These findings highlight the significant impact that road surface conditions can have on traffic-related risks, underscoring the importance of considering these factors in risk management and safety measures.
\newpage

@@

Dark Conditions in (@fig-risk-index-lighting-conditions-1) highlights a concentration of Risk Index values in the moderate-to-high range, indicating that low visibility during dark conditions significantly contributes to elevated collision risks. These findings underscore the importance of improved street lighting or reflective road markers in mitigating risks under dark conditions.

Daylight Conditions distrubition in (@fig-risk-index-lighting-conditions-2) shows a peak at slightly lower Risk Index values compared to dark conditions, reflecting the safety advantages of enhanced visibility. Daylight conditions reduce collision risks, likely due to clear visibility and the predictability of traffic patterns during the day.

@@

Artificial Conditions in (@fig-risk-index-lighting-conditions-4) shows a moderate-to-high concentration of Risk Index values, suggesting that artificial lighting, while helpful, does not fully mitigate risks associated with low visibility. Enhanced lighting technologies and maintenance of artificial lights could further improve safety outcomes under these conditions.

@@

The high collision risk neighborhoods are clustered primarily in specific areas, as denoted by the red points. These regions often correspond to densely populated or high-traffic zones, where the likelihood of traffic collisions is significantly increased. In contrast, the yellow points represent areas where collisions have taken place but with lower collision risks. While these points are more dispersed across the map, they show a notable concentration in downtown Toronto. This clustering in the downtown area suggests that even neighborhoods with moderate or lower collision risks can experience a high frequency of incidents due to the dense urban environment, heavy traffic flows, and increased walkers and cyclist activity typical of central urban areas. This observation highlights the importance of comprehensive urban traffic management strategies across risk levels to improve safety in both high- and low-risk neighborhoods.

@@

The datasets used in the analysis contain missing or incomplete entries, particularly for variables such as road surface conditions and lighting. These gaps may introduce bias or underrepresent specific risk factors, potentially affecting the overall robustness of the findings. Additionally, the Risk Index relies on fixed weights for theft and collision components, which may oversimplify the relative importance of these factors. This static approach might not fully capture variations in risk across different contexts or over time.

@@

Building on the findings and limitations of this study, future research should aim to enhance the understanding of motorbike theft and collision risks through several approaches. One direction is the development of dynamic risk models that integrate real-time data, such as weather conditions, traffic density, and law enforcement activity. These models could provide more immediate and accurate assessments of risk, enabling proactive interventions. Another area for exploration is the impact of seasonal and event-based variations. Examining how risk patterns shift during holidays, festivals, or different seasons would allow for more refined time-sensitive recommendations for both motorbike owners and policymakers.

The inclusion of socioeconomic indicators in future studies is also critical. Variables such as income levels, unemployment rates, and population density could offer a more comprehensive view of the root causes of spatial disparities in risk. Additionally, future research should delve into the interaction effects between environmental and situational factors. For example, investigating how wet roads combined with poor lighting conditions affect risk levels could inform more nuanced interventions.

@@

```{r}
#| label: tbl-raw_data_1
#| tbl-cap: "Examples of raw collision data"
#| echo: false

@@

# Create the example data
mvc_example <- tibble(
  Variable = c(
    "_id", "ACCNUM", "DATE", "TIME", "STREET1", "STREET2", "OFFSET", "ROAD_CLASS",
    "DISTRICT", "ACCLOC", "TRAFFCTL", "VISIBILITY", "LIGHT", "RDSFCOND", 
    "ACCLASS", "IMPACTYPE", "INVTYPE", "INVAGE", "INJURY", "FATAL_NO", 
    "INITDIR", "VEHTYPE", "MANOEUVER", "DRIVACT", "DRIVCOND", "PEDTYPE", 
    "PEDACT", "PEDCOND", "CYCLISTYPE", "CYCACT", "CYCCOND", "PEDESTRIAN", 
    "CYCLIST", "AUTOMOBILE", "MOTORCYCLE", "TRUCK", "TRSN_CITY_VEH", 
    "EMERG_VEH", "PASSENGER", "SPEEDING", "AG_DRIV", "REDLIGHT", "ALCOHOL", 
    "DISABILITY", "HOOD_158", "NEIGHBOURHOOD_158", "HOOD_140", 
    "NEIGHBOURHOOD_140", "DIVISION", "geometry"
  ),
  Example = c(
    "1", "893184", "2006-01-01", "236", "WOODBINE AVE", "O CONNOR DR", "None", 
    "Major Arterial", "Toronto and East York", "Intersection Related", 
    "No Control", "Clear", "Dark", "Wet", "Non-Fatal Injury", "Approaching", 
    "Passenger", "50 to 54", "Major", "None", "None", "None", "None", "None", 
    "None", "None", "None", "None", "None", "None", "None", "None", "Yes", 
    "None", "None", "None", "None", "Yes", "Yes", "Yes", "None", "Yes", "None", 
    "60", "Woodbine-Lumsden", "60", "Woodbine-Lumsden (60)", "D55", 
    '{"type": "MultiPoint", "coordinates": [[-79.318797000266, 43.6995950001932]]}', 
    "N/A"  # Added the missing value
  )
)

# Render the table vertically
mvc_example %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l",
    caption = "Table 1: Example Data from Motor Vehicle Collisions Dataset"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

```{r}
#| label: tbl-raw_data_2
#| tbl-cap: "Examples of raw theft data"
#| echo: false

@@

# Create the example data
example_data <- tibble(
  Variable = c(
    "_id", "EVENT_UNIQUE_ID", "REPORT_DATE", "OCC_DATE", "REPORT_YEAR", 
    "REPORT_MONTH", "REPORT_DAY", "REPORT_DOY", "REPORT_DOW", "REPORT_HOUR", 
    "OCC_YEAR", "OCC_MONTH", "OCC_DAY", "OCC_DOY", "OCC_DOW", "OCC_HOUR", 
    "DIVISION", "LOCATION_TYPE", "PREMISES_TYPE", "UCR_CODE", "UCR_EXT", 
    "OFFENCE", "MCI_CATEGORY", "HOOD_158", "LONG_WGS84", "LAT_WGS84", "geometry"
  ),
  Example = c(
    "1", "GO-20141261501", "2014-01-01", "2014-01-01", "2014", "January", "1", 
    "1", "Wednesday", "8", "2014", "January", "1", "1", "Wednesday", "8", "D51", 
    "Single Home, House (Attach Garage, Cottage, Mobile)", "House", "2142", 
    "200", "Theft From Motor Vehicle Under", "NonMCI", "73", "-79.37453055088850", 
    "43.65706729617110", '{"type": "MultiPoint", "coordinates": [[-79.374530551, 43.657067296]]}'
  )
)

# Render the table vertically
example_data %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l",
    caption = "Table 1: Example Data for Reported Incidents"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

To prepare the raw collision data for analysis, a structured cleaning process was carried out with a focus on maintaining consistency and accuracy. Essential libraries were loaded at the start of the process, including `tidyverse` for data manipulation, `janitor` for standardizing column names, `here` for managing file paths, and `arrow` for saving the cleaned data in an efficient Parquet format. These tools provided a comprehensive framework for handling various cleaning tasks effectively and ensuring the dataset was ready for further use.

@@

In summary, this cleaning process ensured the traffic data was free of duplicates and invalid entries, featured consistent and intuitive column names, and had validated geographic information. The resulting dataset was stored in an optimized format, making it ready for downstream analysis tasks.

```{r}
```

@@

Finally, the cleaned dataset was saved as a Parquet file (`cleaned_crime_data.parquet`) using the `write_parquet()` function. The Parquet format was chosen for its efficiency in storage and processing, making the dataset ready for further analysis. Examples of cleaned theft data is presented in @tbl-cleaned_theft. This comprehensive cleaning process resulted in a high-quality dataset that is well-prepared for any downstream tasks.

```{r}
```

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}

@@

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}

>>>>

---
title: "When Risk Hits the Road: Does Danger Drive a Return?"
subtitle: "Risk Peaks in the Morning on Dark, Snow-Packed Roads and in High Urban Density Areas"
author: 
  - Yingke He
thanks: "Code and data are available at: [https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main](https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main)."
date: today
date-format: long
abstract: "This study examines motor vehicle theft and traffic collisions in Toronto, focusing on spatial and temporal patterns,and environmental factors. A composite risk score model is used and highlights that collisions occur more frequently under unfavorable road conditions, such as poor lighting and wet or icy surfaces, and in areas with inadequate traffic control measures. These findings emphasize the need for targeted interventions, including enhanced road infrastructure and lighting in collision-prone zones. Such measures aim to improve safety and security for motorbike riders and owners, contributing to safer urban mobility in Toronto."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The structure of the paper is organized as follows: following @sec-intro, @sec-data outlines the data collection and preprocessing procedures, along with a detailed description of the outcome variable and the predictor variables used in the analysis. @sec-model introduces the logistic regression models applied to estimate the probability of collision, as well as the method used to derive the theft index and combine these probabilities into a composite risk score. @sec-result then presents the main findings, including understandings into how different factors contribute to the risks of owning and riding a motorbike. Finally, @sec-discussion interprets the results, highlighting significant trends and implications for motorbike risk assessment, and concludes with a discussion on the limitations of the analysis and future research directions.

@@

Details about the data cleaning process and the criteria for variable selection are provided in [Appendix -@sec-data_details].

@@

The Theft from Motor Vehicle dataset contains all reported occurrences of thefts from vehicles, categorized by reported date. These offences are classified based on the value of the stolen items, distinguishing between theft under and theft over thresholds. Each occurrence number may include multiple rows, representing the various offences associated with a single event. The dataset excludes "unfounded" occurrences, adhering to Statistics Canada’s definition that these events were determined not to have occurred or been attempted. Like the KSI dataset, this dataset includes fields for both the old and new neighborhood structures, enabling solid geographic analyses of theft trends.

@@

@fig-risk-index-histogram shows the distribution of the Risk Index across all observations in the dataset. The Risk Index shows a unimodal distribution, skewed slightly to the left, with the majority of values concentrated between 0.45 and 0.65. This indicates that most motorbike-related risks fall within a moderate range. The density curve overlay indicates a smooth progression in risk levels, with the peak occurring around a Risk Index value of 0.55, suggesting that this is the most common level of composite risk. The left tail, representing lower risk levels, is relatively small, while the right tail, corresponding to higher risks, extends further, indicating the presence of a smaller number of high-risk cases.

The skewness and spread of the distribution highlight the variability in the combined risks of theft and collision. The extended tail on the higher end of the Risk Index suggests that certain environmental or situational factors disproportionately increase the risks in specific cases. This idea can inform targeted interventions, focusing on the outliers with high Risk Index values to mitigate the most significant risks.

@@

Building on the distribution of the Risk Index in @fig-risk-index-histogram, @fig-risk-index-boxplot highlights the distribution of the Risk Index across three defined categories: High Risk, Low Risk and Moderate Risk. Each category represents a grouping of neighborhoods based on their average Risk Index values. The plot illustrates distinct differences in the distribution of the Risk Index between these categories.

@@

The **Hour of the Day** variable is used as a predictor to analyze temporal trends in motorbike theft occurrences. This variable helps identify whether certain hours are associated with elevated or reduced theft risks. Contributing factors may include decreased monitoring during nighttime hours, increased activity in high-risk areas during specific times, or patterns related to commuter schedules. @fig-hour_summary outlines the distribution of bicycle theft incidents across different hours of the day in a 24-hour format. This temporal variable helps the model capture time-dependent patterns in theft occurrences, identifying periods of heightened risk, such as late morning to early afternoon (10:00 to 15:00), and lower-risk periods, such as early morning hours (02:00 to 06:00).By including this variable, the theft index model utilizes the observed trend of thefts being more frequent during morning and early afternoon hours (7:00 to 13:00) while being significantly lower during the early morning hours (02:00 to 06:00). This temporal variability highlights specific periods of increased theft risk, enabling the model to capture these patterns effectively. Understanding this trend allows for better predictions of theft dynamics and offers meaningful ideas for designing targeted interventions and optimizing resource allocation based on the time of day.

@@

The Collision Probability model, on the other hand, examines situational and environmental conditions affecting the likelihood of collisions. Variables such as Neighborhood ID, Road Surface Condition, Lighting Condition, Traffic Control, Time of Day, and Visibility provides an understanding into the contextual factors that contribute to traffic incidents, capturing the dynamic interaction between environmental factors and human behavior.

@@

The **Traffic Control** variable categorizes the type of traffic management present at the site of each incident, such as "No Control," "Traffic Signal," or "Stop Sign." This variable provides ideas into how various traffic control measures influence the likelihood of motorbike collisions.

@@

@@

This modeling approach is justified by its ability to adapt to the data’s constraints while maintaining statistical rigor and interpretability. The sub-index method for theft risk is tailored to the dataset’s characteristics, avoiding assumptions about unobserved cases, and the use of logistic regression for collision risk ensures robust and reliable predictions. Together, these components form a practical and practical framework for evaluating motorbike ownership and usage risks, addressing both immediate safety concerns and long-term theft risks.

@@

where P represents the predicted probability of collision from the logistic regression model plotted against the continuous predictors. These plots are intended to visualize the relationship between each predictor and the logit of the outcome variable, providing thoughts into whether the relationship is approximately linear.

@@

Despite this overlap, both predictors are retained in the model because they provide distinct and meaningful contributions to understanding collision risk. Visibility Conditions directly reflects environmental factors like fog, heavy rain, or low light, which impair drivers' ability to see hazards. Conversely, Road Conditions\` account for the physical state of the driving surface, such as wet, icy, or damaged roads, which influence vehicle handling and stopping distance. Together, these variables capture complementary aspects of collision risk, ensuring that the model provides a solid assessment.

@@

The year 2020 coincides with the beginning of the COVID-19 pandemic, which had a strong impact on urban mobility patterns. Lockdowns, social distancing measures, and reduced economic activities likely led to significant decreases in traffic volumes and changes in behavior related to motor vehicle usage [@buehler2021covid]. However, the Risk Index remains relatively consistent throughout the year, with only minor fluctuations across months. This relative stability may suggest that while pandemic-related restrictions likely influence traffic and mobility patterns, these changes did not significantly impact the overall Risk Index for the year. The consistent trends could reflect a balance between reductions in traffic collisions due to decreased activity during lockdown periods and a stable baseline for theft and other risk factors.

@@

This section examines how factors such as road surface conditions, lighting conditions, and traffic control types contribute to changes in the Risk Index. Categorizing the data based on these factors illustrates specific conditions under which the likelihood of collisions or theft increases.

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing understandings into how these factors may contribute to the overall risk.
```{r}
#| label: fig-risk-distribution-roadcondition
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Road Surface Conditions"
#| fig-pos: H
#| fig-width: 8.5
#| fig-height: 4

@@

@fig-risk-distribution-roadcondition highlights notable differences in the Risk Index distribution across various road surface conditions. Wet and icy conditions, in particular, show higher variability in the Risk Index, suggesting that these surfaces may lead to more unpredictable and elevated risk levels. Dry conditions, on the other hand, generally exhibit a lower and more consistent risk. These findings highlight the significant impact that road surface conditions can have on traffic-related risks, underscoring the importance of considering these factors in risk management and safety measures.
\newpage

@@

Dark Conditions in @fig-risk-index-lighting-conditions-1 highlights a concentration of Risk Index values in the moderate-to-high range, indicating that low visibility during dark conditions significantly contributes to elevated collision risks. These findings underscore the importance of improved street lighting or reflective road markers in mitigating risks under dark conditions.

Daylight Conditions distrubition in @fig-risk-index-lighting-conditions-2 shows a peak at slightly lower Risk Index values compared to dark conditions, reflecting the safety advantages of enhanced visibility. Daylight conditions reduce collision risks, likely due to clear visibility and the predictability of traffic patterns during the day.

@@

Artificial Conditions in @fig-risk-index-lighting-conditions-4 shows a moderate-to-high concentration of Risk Index values, suggesting that artificial lighting, while helpful, does not fully mitigate risks associated with low visibility. Enhanced lighting technologies and maintenance of artificial lights could further improve safety outcomes under these conditions.

@@

The high collision risk neighborhoods are clustered primarily in specific areas, as denoted by the red points. These regions often correspond to densely populated or high-traffic zones, where the likelihood of traffic collisions is significantly increased. In contrast, the yellow points represent areas where collisions have taken place but with lower collision risks. While these points are more dispersed across the map, they show a notable concentration in downtown Toronto. This clustering in the downtown area suggests that even neighborhoods with moderate or lower collision risks can experience a high frequency of incidents due to the dense urban environment, heavy traffic flows, and increased walkers and cyclist activity typical of central urban areas. This observation highlights the importance of urban traffic management strategies across risk levels to improve safety in both high- and low-risk neighborhoods.

@@

The datasets used in the analysis contain missing or incomplete entries, particularly for variables such as road surface conditions and lighting. These limitations may introduce bias or underrepresent specific risk factors, potentially affecting the overall realiability of the findings. Additionally, the Risk Index relies on fixed weights for theft and collision components, which may oversimplify the relative importance of these factors. This static approach might not fully capture variations in risk across different contexts or over time.

@@

Building on the findings and limitations of this study, future research should aim to enhance the understanding of motorbike theft and collision risks through several approaches. One direction is the development of dynamic risk models that integrate real-time data, such as weather conditions, traffic density, and law enforcement activity. These models could provide more immediate and accurate assessments of risk, enabling proactive interventions. Another area for research is the impact of seasonal and event-based variations. Examining how risk patterns shift during holidays, festivals, or different seasons would allow for more refined time-sensitive recommendations for both motorbike owners and policymakers.

The inclusion of socioeconomic indicators in future studies is also critical. Variables such as income levels, unemployment rates, and population density could offer a more thorough view of the root causes of spatial disparities in risk. Additionally, future research should explore the interaction effects between environmental and situational factors. For example, investigating how wet roads combined with poor lighting conditions affect risk levels could inform more nuanced interventions.

@@

```{r}
#| label: tbl-raw_data_1
#| tbl-cap: "Examples of Raw Collision Data"
#| echo: false

@@

# Create the example data
mvc_example <- tibble(
  Variable = c(
    "_id", "ACCNUM", "DATE", "TIME", "STREET1", "STREET2", "OFFSET", "ROAD_CLASS",
    "DISTRICT", "ACCLOC", "TRAFFCTL", "VISIBILITY", "LIGHT", "RDSFCOND", 
    "ACCLASS", "IMPACTYPE", "INVTYPE", "INVAGE", "INJURY", "FATAL_NO", 
    "INITDIR", "VEHTYPE", "MANOEUVER", "DRIVACT", "DRIVCOND", "PEDTYPE", 
    "PEDACT", "PEDCOND", "CYCLISTYPE", "CYCACT", "CYCCOND", "PEDESTRIAN", 
    "CYCLIST", "AUTOMOBILE", "MOTORCYCLE", "TRUCK", "TRSN_CITY_VEH", 
    "EMERG_VEH", "PASSENGER", "SPEEDING", "AG_DRIV", "REDLIGHT", "ALCOHOL", 
    "DISABILITY", "HOOD_158", "NEIGHBOURHOOD_158", "HOOD_140", 
    "NEIGHBOURHOOD_140", "DIVISION", "geometry"
  ),
  Example = c(
    "1", "893184", "2006-01-01", "236", "WOODBINE AVE", "O CONNOR DR", "None", 
    "Major Arterial", "Toronto and East York", "Intersection Related", 
    "No Control", "Clear", "Dark", "Wet", "Non-Fatal Injury", "Approaching", 
    "Passenger", "50 to 54", "Major", "None", "None", "None", "None", "None", 
    "None", "None", "None", "None", "None", "None", "None", "None", "Yes", 
    "None", "None", "None", "None", "Yes", "Yes", "Yes", "None", "Yes", "None", 
    "60", "Woodbine-Lumsden", "60", "Woodbine-Lumsden (60)", "D55", 
    '{"type": "MultiPoint", "coordinates": [[-79.318797, 43.699595]]}', 
    "N/A"  # Added the missing value
  )
)

# Render the table vertically
mvc_example %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l",
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

```{r}
#| label: tbl-raw_data_2
#| tbl-cap: "Examples of Raw Theft Data"
#| echo: false

@@

# Create the example data
example_data <- tibble(
  Variable = c(
    "_id", "EVENT_UNIQUE_ID", "REPORT_DATE", "OCC_DATE", "REPORT_YEAR", 
    "REPORT_MONTH", "REPORT_DAY", "REPORT_DOY", "REPORT_DOW", "REPORT_HOUR", 
    "OCC_YEAR", "OCC_MONTH", "OCC_DAY", "OCC_DOY", "OCC_DOW", "OCC_HOUR", 
    "DIVISION", "LOCATION_TYPE", "PREMISES_TYPE", "UCR_CODE", "UCR_EXT", 
    "OFFENCE", "MCI_CATEGORY", "HOOD_158", "LONG_WGS84", "LAT_WGS84", "geometry"
  ),
  Example = c(
    "1", "GO-20141261501", "2014-01-01", "2014-01-01", "2014", "January", "1", 
    "1", "Wednesday", "8", "2014", "January", "1", "1", "Wednesday", "8", "D51", 
    "Single Home, House (Attach Garage, Cottage, Mobile)", "House", "2142", 
    "200", "Theft From Motor Vehicle Under", "NonMCI", "73", "-79.37453055088850", 
    "43.65706729617110", '{"type": "MultiPoint", "coordinates": [[-79.3745305, 43.6570672]]}'
  )
)

# Render the table vertically
example_data %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

@@

To prepare the raw collision data for analysis, a structured cleaning process was carried out with a focus on maintaining consistency and accuracy. Essential libraries were loaded at the start of the process, including `tidyverse` for data manipulation, `janitor` for standardizing column names, `here` for managing file paths, and `arrow` for saving the cleaned data in an efficient Parquet format. These tools provided a solid framework for handling various cleaning tasks effectively and ensuring the dataset was ready for further use.

@@

In summary, this cleaning process ensured the collision data was free of duplicates and invalid entries, featured consistent and intuitive column names, and had validated geographic information. The resulting cleaned dataset show in @tbl-cleaned_collision_data was stored in an optimized format, making it ready for downstream analysis tasks.

```{r}
#| label: tbl-cleaned_collision_data
#| tbl-cap: "Example of Cleaned Collision Data"
#| echo: false

# Create the example data
collision_data_example <- tibble(
  Variable = c(
    "id", "accident_id", "report_date", "time_of_day", "street_primary", 
    "street_secondary", "offset", "road_type", "district", "accident_location", 
    "traffic_control", "visibility_conditions", "lighting_conditions", 
    "road_conditions", "accident_classification", "impact_type", "invtype", 
    "invage", "injury", "fatal_no", "initdir", "vehicle_type", "maneuver", 
    "driver_action", "driver_condition", "pedtype", "pedact", "pedcond", 
    "cyclistype", "cycact", "cyccond", "pedestrian", "cyclist", "automobile", 
    "motorcycle", "truck", "trsn_city_veh", "emerg_veh", "passenger", "speeding", 
    "ag_driv", "redlight", "alcohol", "disability", "hood_158", "neighborhood", 
    "hood_140", "neighbourhood_140", "division"
  ),
  Example = c(
    "1", "893184", "2006-01-01", "236", "WOODBINE AVE", "O CONNOR DR", "NA", 
    "Major Arterial", "Toronto and East York", "Intersection Related", 
    "No Control", "Clear", "Dark", "Wet", "Non-Fatal Injury", "Approaching", 
    "Passenger", "NA", "Major", "NA", "NA", "NA", "NA", "NA", "NA", "NA", 
    "NA", "NA", "NA", "NA", "NA", "NA", "NA", "Yes", "NA", "NA", "NA", "NA", 
    "Yes", "Yes", "Yes", "NA", "Yes", "NA", "60", "Woodbine-Lumsden", "60", 
    "Woodbine-Lumsden (60)", "D55"
  )
)

# Render the table vertically
collision_data_example %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l",
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  ) %>%
  column_spec(2, width = "20em")  # Adjust width to handle longer text

@@

Finally, the cleaned dataset was saved as a Parquet file (`cleaned_crime_data.parquet`) using the `write_parquet()` function. The Parquet format was chosen for its efficiency in storage and processing, making the dataset ready for further analysis. Examples of cleaned theft data is presented in @tbl-cleaned_theft. This cleaning process resulted in a high-quality dataset that is well-prepared for any downstream tasks.

```{r}
#| label: tbl-cleaned_theft
#| tbl-cap: "Example of Cleaned Theft Data"
#| echo: false

# Create the example data
cleaned_data_example <- tibble(
  Variable = c(
    "event_id", "report", "occurrence", "hour", "day_of_week", "division",
    "location", "premises_type", "offense", "mci_category", "hood_158", 
    "longitude", "latitude"
  ),
  Example = c(
    "GO-20141261501", "2014-01-01", "2014-01-01", "8", "Wednesday", "D51", 
    "Single Home, House (Attach Garage, Cottage, Mobile)", "House", 
    "Theft From Motor Vehicle Under", "NonMCI", "73", "-79.37453055088850", 
    "43.65706729617110"
  )
)

# Render the table vertically
cleaned_data_example %>%
  kable(
    col.names = c("Variable Name", "Example Value"),
    align = "l"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  ) %>%
  column_spec(2, width = "20em")  # Adjust width for better text wrapping

@@

## Predictor Variable Choice

### Theft Index

In constructing the theft risk sub-indexes, the choice of predictor variables: Hour of the Day, Day of the Week, and Premises Type was informed by both theoretical and practical considerations rooted in spatial, temporal, and environmental factors associated with theft incidents. These variables were selected from the cleaned theft dataset to create a strong and interpretable model that captures key patterns of theft occurrences while aligning with the study's goal of identifying targeted interventions.

1. Hour of the Day

The inclusion of the Hour of the Day as a predictor variable reflects the strong temporal dependency of theft occurrences. Theft risk often varies significantly across a 24-hour period due to human activity patterns, such as peak commuting times, work hours, and late-night inactivity. For example, motor vehicle thefts are more likely to occur during nighttime hours when there is reduced surveillance and fewer bystanders to deter criminal behavior. By incorporating the Hour of the Day into the theft risk sub-index, the model accounts for this temporal dimension, enabling stakeholders to identify high-risk hours and implement time-specific interventions, such as increased nighttime patrols or surveillance in parking lots and residential areas.

2. Day of the Week
